---
title: "CMPT318-A2"
author: "Group 10"
date: "October 26, 2018"
output: html_document
---

```{r include=FALSE}
library("depmixS4")
library(lubridate)

Dataset1 <- read.table("./Dataset.txt", header=TRUE, sep=",") #525600 with NA values
#Dataset2 <- na.omit(Dataset2) #521860 with na values omitted.

# Set the Sunday mornings and nights
dates <- as.POSIXlt(Dataset1$Date, na.rm = T, format = "%d/%m/%Y") # get date format
Dataset1$Day <- weekdays(as.Date(dates)) # add corresponding days to the dataset
sunday <- Dataset1$Day == "Sunday"
monday <- Dataset1$Day == "Monday"
times <- format(Dataset1$Time, na.rm = T, format = "%H:%M:%S")
sundayMornings <- subset(Dataset1, sunday & times >= "08:00:00" & times <= "11:00:00")
sundayNights <- subset(Dataset1, (sunday & times >= "21:00:00" & times <= "23:59:00") | (monday & times == "00:00:00"))
sundayNights <- sundayNights[-1,]
```

# Question 1
```{r}
# Question 1
# Set the variables
GAP <- sundayMornings$Global_active_power
GRP <- sundayMornings$Global_reactive_power
Vol <- sundayMornings$Voltage
GI  <- sundayMornings$Global_intensity

set.seed(1)
BICvector <- vector()
logLikVector <- vector()
states <- c(2:10)
vectorntimes = rep(181, nrow(sundayMornings)/181)
for (state in states) {
  print(paste("Number of States: ", state))
  mod <- depmix(response = Global_active_power ~ 1, data = sundayMornings, nstates = state, ntimes = vectorntimes)
  fm <- fit(mod)
  summary(fm)
  BICvector <- c(BICvector, BIC(fm))
  logLikVector <- c(logLikVector, logLik(fm))
}
plot(states, BICvector, type="o", col="red", main="Hidden Markov Models (Sunday Mornings)", xlab = "Number of States", ylab = "BIC Value of Model")
plot(states, logLikVector, type="o", col="blue", main="Hidden Markov Models (Sunday Mornings)", xlab = "Number of States", ylab = "Log Likelihood of Model")
```

Results:
We split the data set into the time frame of Sunday Mornings which are 8 AM to 11 AM (inclusive). Once the we have the dataset to perform the analysis on, we run the depmix function with 2 to 10 states and plot the results on a graph with the respective BIC values and the negative Log Likelihood values. 
From the plots we have obtained, we see that the ideal number of states for the sundayMorning data set is 8. The reason we chose 8 states is because the BIC value goes up with 9 states. Any number of states higher than 8 seem to be overfitting the model.
The exact BIC value when the model has 8 states is: 5833.805
The exact Log Likelihood when the model has 8 states is: -2555.488

# Question 2
```{r}
# Question 2

set.seed(1)
BICvector <- vector()
logLikVector <- vector()
states <- c(2:15)
vectorntimes = rep(181, nrow(sundayNights)/181)

for (state in states) {
  print(paste("Number of States: ", state))
  mod <- depmix(response = Global_active_power ~ 1, data = sundayNights, nstates = state, ntimes = vectorntimes)
  fm <- fit(mod)
  summary(fm)
  BICvector <- c(BICvector, BIC(fm))
  logLikVector <- c(logLikVector, logLik(fm))
}

plot(states, BICvector, type="b", col="red", main="Hidden Markov Models (Sunday Nights)", xlab = "Number of States", ylab = "BIC Value of Model")
plot(states, logLikVector, type="b", col="blue", main="Hidden Markov Models (Sunday Nights)", xlab = "Number of States", ylab = "Log Likelihood of Model")

```
Results:
We split the data set into the time frame of Sunday Nights which are 9 PM to 12 AM (inclusive). Once we have the dataset to perform the analysis on, we run the depmix function with 2 to 15 states and plot the results on a graph with the respective BIC values and the Log Likelihood values. 
From the plots we have obtained, we see that the ideal number of states for the Sunday Nights data set is 10. The reason we chose 10 states is because the BIC value goes up with 11 states. Any number of states higher than 10 seem to be overfitting the model. 
The exact BIC value when the model has 10 states is: 9133.780
The exact Log Likelihood when the model has 10 states is: -4022.480


# Question 3 part a 
```{r}
c <- 0
avgGAPMorningVector <- vector()
morningTime <- unique(format(sundayMornings$Time, na.rm = T, format = "%H:%M:%S"))
for (time in morningTime) {
  c <- c + 1
  avgGAPMorning <- mean(sundayMornings$Global_active_power[times == time], na.rm = T)
  avgGAPMorningVector <- c(avgGAPMorningVector, avgGAPMorning)
  print(paste(c, "Average Global_active_power for Sunday mornings at time", time, "=", avgGAPMorning))
}
c <- 0
avgGAPNightVector <- vector()
nightTime <- unique(format(sundayNights$Time, na.rm = T, format = "%H:%M:%S"))
for (time in nightTime) {
  c <- c + 1
  avgGAPNight <- mean(sundayNights$Global_active_power[times == time], na.rm = T)
  avgGAPNightVector <- c(avgGAPNightVector, avgGAPNight)
  print(paste(c, "Average Global_active_power for Sunday nights at time", time, "=", avgGAPNight))
}

# Plot Average Global_active_power for Sunday mornings and nights
morningTime <- strptime(morningTime, format = "%H:%M:%S")
plot(morningTime, avgGAPMorningVector, ty="l", col="red")
nightTime <- factor(nightTime, levels=sundayNights[0:181,]$Time)
plot(nightTime, avgGAPNightVector, type="n", col="blue", xlab="Sunday Night Times", ylab="Average Global Active Power")
lines(nightTime, avgGAPNightVector, type="l", col="blue", xlab="Sunday Night Times", ylab="Average Global Active Power")

# Plot Time Series for average Global_active_power for Sunday mornings and nights
avgGAPMorningTS <- ts(avgGAPMorningVector)
avgGAPNightTS <- ts(avgGAPNightVector)
plot.ts(avgGAPMorningTS, col="red")
plot.ts(avgGAPNightTS, col="blue")

```

Explanation:
The red plots represent the average global active power plotted against morning times. The blue plots are for Sunday night times. We see that from 9:30 PM to 11 PM the most electricity was consumed.
We see that for the Sunday morning times the electricity consumption is high from 9:30 AM to 11 AM.


# Question 3 part b
```{r}
sundaydate <- unique(sundayMornings$Date)

# Defining weekly statistical measurements
weekeveningmin <- vector()
weekeveningmax <- vector()
weekeveningavg <- vector()
weekeveningsd  <- vector()  
weekmorningmin <- vector()
weekmorningmax <- vector()
weekmorningavg <- vector()
weekmorningsd  <- vector()  

# Defining monthly statistical measurements
monthmorningmin <- vector()
monthmorningmax <- vector()
monthmorningavg <- vector()
monthmorningsd  <- vector()
montheveningmin <- vector()
montheveningmax <- vector()
montheveningavg <- vector()
montheveningsd  <- vector() 

# Defining seasonal statistical measurements
seasonmorningmin <- vector()
seasonmorningmax <- vector()
seasonmorningavg <- vector()
seasonmorningsd  <- vector()  
seasoneveningmin <- vector()
seasoneveningmax <- vector()
seasoneveningavg <- vector()
seasoneveningsd  <- vector()  

i = 0
for (Date in sundaydate){
  i <- i + 1
  minlist <- sort(sundayMornings$Global_active_power[which(sundayMornings$Date == Date)])
  maxlist <- sort(minlist,decreasing = T)
  weekmorningmin[i] <- minlist[1]
  weekmorningmax[i] <- maxlist[1]
  weekmorningavg[i] <- mean(minlist)
  weekmorningsd[i]  <- sd(minlist)
  
  minlist <- sort(sundayNights$Global_active_power[((181*i)-180):(181*i)])
  maxlist <- sort(minlist,decreasing = T)
  weekeveningmin[i] <- minlist[1]
  weekeveningmax[i] <- maxlist[1]
  weekeveningavg[i] <- mean(minlist)
  weekeveningsd[i]  <- sd(minlist)
}

months <- (1:12)

for (i in 1:12){
  targetmon <- subset(sundayMornings, month(sundayMornings$Date) == i)
  minlist <- sort(targetmon$Global_active_power)
  maxlist <- sort(minlist,decreasing = T)
  monthmorningmin[i] <- minlist[1]
  monthmorningmax[i] <- maxlist[1]
  monthmorningavg[i] <- mean(minlist)
  monthmorningsd[i]  <- sd(minlist)
  
  targetmon <- subset(sundayNights, month(sundayNights$Date) == i)
  minlist <- sort(targetmon$Global_active_power)
  maxlist <- sort(minlist,decreasing = T)
  montheveningmin[i] <- minlist[1]
  montheveningmax[i] <- maxlist[1]
  montheveningavg[i] <- mean(minlist)
  montheveningsd[i]  <- sd(minlist)
}


for (i in 1:4){
  
  if (((3*i)+2) < 12){
    targetseason <- subset(sundayMornings, month(sundayMornings$Date) >= (3*i) & month(sundayMornings$Date) <= ((3*i)+2))
    minlist <- sort(targetseason$Global_active_power)
    maxlist <- sort(minlist,decreasing = T)
    seasonmorningmin[i] <- minlist[1]
    seasonmorningmax[i] <- maxlist[1]
    seasonmorningavg[i] <- mean(minlist)
    seasonmorningsd[i]  <- sd(minlist)
    
    targetseason <- subset(sundayNights, month(sundayNights$Date) >= (3*i) & month(sundayNights$Date) <= ((3*i)+2))
    minlist <- sort(targetseason$Global_active_power)
    maxlist <- sort(minlist,decreasing = T)
    seasoneveningmin[i] <- minlist[1]
    seasoneveningmax[i] <- maxlist[1]
    seasoneveningavg[i] <- mean(minlist)
    seasoneveningsd[i]  <- sd(minlist)
  }else{
    targetseason <- subset(sundayMornings, month(sundayMornings$Date) == 1 | month(sundayMornings$Date) == 12 | month(sundayMornings$Date) == 2)
    minlist <- sort(targetseason$Global_active_power)
    maxlist <- sort(minlist,decreasing = T)
    seasonmorningmin[i] <- minlist[1]
    seasonmorningmax[i] <- maxlist[1]
    seasonmorningavg[i] <- mean(minlist)
    seasonmorningsd[i]  <- sd(minlist)
    
    targetseason <- subset(sundayNights, month(sundayNights$Date) == 1 | month(sundayNights$Date) == 12 | month(sundayNights$Date) == 2)
    minlist <- sort(targetseason$Global_active_power)
    maxlist <- sort(minlist,decreasing = T)
    seasoneveningmin[i] <- minlist[1]
    seasoneveningmax[i] <- maxlist[1]
    seasoneveningavg[i] <- mean(minlist)
    seasoneveningsd[i]  <- sd(minlist)
  }
}

# Weekly
plot(1:52,weekmorningmin, xlab ='Week', ylab = 'Weekly Min on Sunday Morning', type = "o")
plot(1:52,weekmorningmax, xlab ='Week', ylab = 'Weekly Max on Sunday Morning', type = "o")
plot(1:52,weekmorningavg, xlab ='Week', ylab = 'Weekly Avg on Sunday Morning', type = "o")
plot(1:52,weekmorningsd,  xlab ='Week', ylab = 'Weekly Sd on Sunday Morning',  type = "o")
plot(1:52,weekeveningmin, xlab ='Week', ylab = 'Weekly Min on Sunday Evening', type = "o")
plot(1:52,weekeveningmax, xlab ='Week', ylab = 'Weekly Max on Sunday Evening', type = "o")
plot(1:52,weekeveningavg, xlab ='Week', ylab = 'Weekly Avg on Sunday Evening', type = "o")
plot(1:52,weekeveningsd,  xlab ='Week', ylab = 'Weekly Sd on Sunday Evening',  type = "o")

# Monthly
plot(1:12,monthmorningmin, xlab ='Month', ylab = 'Monthly Min on Sunday Morning', type = "o")
plot(1:12,monthmorningmax, xlab ='Month', ylab = 'Monthly Max on Sunday Morning', type = "o")
plot(1:12,monthmorningavg, xlab ='Month', ylab = 'Monthly Avg on Sunday Morning', type = "o")
plot(1:12,monthmorningsd,  xlab ='Month', ylab = 'Monthly Sd on Sunday Morning',  type = "o")
plot(1:12,montheveningmin, xlab ='Month', ylab = 'Monthly Min on Sunday Evening', type = "o")
plot(1:12,montheveningmax, xlab ='Month', ylab = 'Monthly Max on Sunday Evening', type = "o")
plot(1:12,montheveningavg, xlab ='Month', ylab = 'Monthly Avg on Sunday Evening', type = "o")
plot(1:12,montheveningsd,  xlab ='Month', ylab = 'Monthly Sd on Sunday Evening',  type = "o")

# Seasonal
plot(xlim=c(1,5),seasonmorningmin, xlab ='Season', ylab = 'Seasonal Min on Sunday Morning', type = "o")
plot(xlim=c(1,5),seasonmorningmax, xlab ='Season', ylab = 'Seasonal Max on Sunday Morning', type = "o")
plot(xlim=c(1,5),seasonmorningavg, xlab ='Season', ylab = 'Seasonal Avg on Sunday Morning', type = "o")
plot(xlim=c(1,5),seasonmorningsd,  xlab ='Season', ylab = 'Seasonal Sd on Sunday Morning',  type = "o")
plot(xlim=c(1,5),seasoneveningmin, xlab ='Season', ylab = 'Seasonal Min on Sunday Evening', type = "o")
plot(xlim=c(1,5),seasoneveningmax, xlab ='Season', ylab = 'Seasonal Max on Sunday Evening', type = "o")
plot(xlim=c(1,5),seasoneveningavg, xlab ='Season', ylab = 'Seasonal Avg on Sunday Evening', type = "o")
plot(xlim=c(1,5),seasoneveningsd,  xlab ='Season', ylab = 'Seasonal Sd on Sunday Evening',  type = "o")

```
Explanation:
Season 1 - Spring
Season 2 - Summer
Season 3 - Fall
Season 4 - Winter

For seasonal plots, we can see that in Winter the SD in Global Electric Power consumption is high. The electricity deviates largely in Winter.
Looking at the plots for mean, minimum, maximum and standard deviation, we have come to the conclusion that electricity consumption is high during Winter and the lowest in the Summer. Looking at a broader picture allows us to see more general trends of electricity consumption.